---
title: "R Notebook - Lesson 6 Homework Template"
output: html_notebook
---

Lesson 6 Homework:

Part I: CSV to Excel
Part II: Wrangling the language_diversity data set

NOTE: To install the "untidydata" package and obtain the language_diversity data set, type the following into your R console: devtools::install_github("jvcasillas/untidydata")

```{r}
library(tidyverse)
library(xlsx)
library(untidydata)
```

Part I: CSV to Excel
Read in the P2-movie-ratings.csv file
Look at how the data types were parsed either before or after you read the file into a tibble
Revise the file data types as follows:
     - Film, chr
     - Genre, factor / category
     - Ratings %s, Budget and Year of Release, int

Call the summary function on the int columns
Determine the 75th percentile for the Audience Rating %, and create a new column, `Highly Rated`, which is 1 when Audience Rating % is > 75th percentile, and 0 when it isn't.
Create a summary by Genre showing the Average Ratings and the number of Highly Rated films in each category
Arrange the summary data into a pleasing format and rename the colnames so they are easy to understand
Convert the revised tibble and summary to data frames
Write the data frames to an Excel document, 'P2-movie-ratings.xlsx' as Sheet 1 and Sheet 2


```{r}
movies <- read_csv("P2-movie-ratings.csv", col_names=TRUE)

spec(movies)
```

Part II: Wrangling the language_diversity data set

Install the ‘untidydata’ library using the following command: devtools::install_github("jvcasillas/untidydata")
Tidy the language_diversity data set by spreading the data in the Measurement and Value columns into the following columns: Languages, Area, Population, Stations, MGS, and Std
Create a new tibble by Selecting columns except MGS and Std
Cast the Continent column as a factor
Filter the tibble to only Asian countries
Mutate to create a new column, `Language Rate` that contains languages per area
Group the data by country and arrange it by Area
Summarize each country by min, mean and max of `Language Rate` to see which countries in Asia have the most languages per area
Write the new summarized data set to a .csv file using write_csv (the readr function that works with tibbles)


```{r}
language_diversity
```

