---
title: "R Notebook: Lesson 4, Vectors & Matrices"
output: html_notebook
---

Defining a Function

```{r}

# A function that takes zero arguments and returns nothing

foo <- function() {
  print("This is my first function, foo.")
}

foo()

# A function that takes 1 argument, and returns nothing

foo <- function(x) {
  print(paste("You passed ", x, "to my second function, foo."))
}

foo(88)
foo("Dweezl")

# A function that takes 2 arguments and returns one value
foo <- function(x, y=0) {
  print(paste("You passed ", x, " and ", y, " to my third function, foo."))
  return(x*y)
}

foo(88)
foo(11, 12)
foo(5, 7)

```
Functions and Scope

```{r}

MY_SPEED_LIMIT <<- 110

# GLOBAL x
x = 15
c = 22
b = 90

foo <- function(x) {
  # LOCAL x, x that was passed to the function foo
  print(paste("The value of x is ", x))
}


foo(b)
foo(c)

# Once foo exits, what is the value of x?
print(paste("The value of x is ", x))

```

A Recursive Function

A function that calls itself is called a recursive function.
A recursive function must contain a terminating condition.

Here's an example of a recursive function that terminates when the value passed to it becomes less than or equal to 1

```{r}
# Program to convert decimal number into binary number using recursive function
convert_to_binary <- function(n) {
  
  print(paste("n is ", n))
  if(n > 1) {
    convert_to_binary(as.integer(n/2))
  }
  cat(n %% 2)
}

convert_to_binary(54)
```
```{r}
convert_to_binary(101)

```

Over-riding an Infix Operator
In R, it is possible - for example - to define your own version of +, if you surround the operator with back-ticks (above the Tab key)

```{r}
# a function that adds strings together
`+` <- function(a,b) {
 return(paste(a, b))
}

"Hello" + "World!"

x = "The artist is..."
y = "Claude Monet"
x + y

print("I love those haystacks by" + y)


```



