---
title: 'R Notebook: Lesson 4, Data Structure & Variable Review'
output:
  html_document:
    df_print: paged
---

All variables are objects!

Good & Not-so-good names for variables / objects

A name cannot start with a number. 

A name cannot use some special symbols, like ^, !, $, @, +, -, /, or *

```{r}
a <- c("one", "two", "three")
a

VAULTS <- c("Tsukahara", "Yurchenko")
VAULTS
```


```{r}
1VAULTS <- c("Tsukahara", "Yurchenko")

print(1VAULTS)
```

Don't forget that R is case-sensitive:

```{r}
meany <- mean(c(-1, 2, 3, 1, 0, 1, 0, -2, 3, 4))
Meany <- "not so nice, there."

meany
Meany
```

Data structures

Homogeneous: 1d, Atomic vector
             2d, Matrix
             nd, Array

Heterogeneous: List, Data Frame

In precalculus, we all received an introduction to the difference between scalars (measures of distance, or magnitude), and vectors (measures of distance / magnitude AND direction).  R has no scalars.  Vectors are so essential to R, that even its scalars are vectors of *unit length*.


```{r}
# Vectors

i_am_a_vector <- 9
me_too <- 1:9

i_am_a_vector
me_too

me_too + i_am_a_vector
me_too * i_am_a_vector

# VAULTS is a character vector of size 2
VAULTS

```

```{r}

and_so_am_i <- 1:10
and_so_am_i
i_am_a_vector * and_so_am_i

```

Dress Shopping with 'next' and 'break'.

```{r}
dresses <- c("Prada", "CK", "YSL", "Gucci", "Guess")

for(d in dresses) {
  
  print(d)

  if (d == "CK") {
    print("I don’t like CK! Please bring me the next one.")
	  next
  }
  
  if (d == "Gucci") {
    print("Squee! I found it!")
    break
  }

}

```

Lesser Used: The repeat loop

```{r}
x <- 10
repeat {
  print(paste(x, " bottles of beer on the wall, ", x, " bottles of beer!"))

  if (x == 1){
    print("Saving the last one for later!")
    break
  } else {
    print("Take one down and pass it around...")
  }
  x = x-1
}
```




